\ProjectName{}: A Database for Missing Data

### 2016-12-12 paper/algorithms/histogram-transform.tex

\begin{algorithm}
  \begin{algorithmic}
    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}

    \Function{UpdateHistogram}{$H, op$}
    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
		\Return{H}
	\EndIf
	\State $H' \gets \Call{Copy}{H}$
    	\If{op = $\delta_C$}
		\For{$c \in C$}
			\State $H'[c][null] \gets 0$
		\EndFor
	\ElsIf{op = $\mu_C$}
		\For{$c \in C$}
			\State \Call{Add}{$H'[c], H'[c][null]$}
		\EndFor
	\ElsIf{op = $\sigma$}
		\State \Call{ScaleBy}{H',  \Call{Selectivity}{$\sigma$}}
	\ElsIf{op = $\bowtie_\phi$}
		\State \Call{ScaleTo}{H', \Call{Cardinality}{$\bowtie_{\phi}$}}
	\EndIf
	\Return{H'}
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for in-plan histogram updates}
  \label{algo:histogram-transformation}
\end{algorithm}

>>>>

%\begin{algorithm}
  \begin{algorithmic}
    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}

    \Function{UpdateHistogram}{$H, op$}
    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
		\Return{H}
	\EndIf
	\State $H' \gets \Call{Copy}{H}$
    	\If{op = $\delta_C$}
		\For{$c \in C$}
			\State $H'[c][null] \gets 0$
		\EndFor
	\ElsIf{op = $\mu_C$}
		\For{$c \in C$}
			\State \Call{Add}{$H'[c], H'[c][null]$} 	\Comment{Adds to histogram based on distirbution}
		\EndFor
	\ElsIf{op = $\sigma$}
		\State \Call{ScaleBy}{$H',  \Call{Selectivity}{\sigma}$} 	\Comment{Scales all buckets by a constant factor}
	\ElsIf{op = $\bowtie_\phi$}
		\State \Call{ScaleTo}{$H', \Call{Cardinality}{\bowtie_{\phi}}$} 	\Comment{Scales all buckets to sum to constant}
	\EndIf
	\Return{H'}
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for in-plan histogram updates}
  \label{algo:histogram-transformation}
%\end{algorithm}

### 2016-12-12 paper/algorithms/plan-helpers.tex

\begin{algorithm}
  \begin{algorithmic}
    \Require{$q$ is a query plan, $C_{req}$ is a set of attributes that must be imputed.}
    \Ensure{Returns a set of query plans such that $\Call{Dirty}{q'} \cap C_{req} = \emptyset$.}
    \Function{AddImpute}{$q, C_{req}$}
    \todo{jose: I modified this, since in reality we return only the same plan if it has no dirty values, we may still impute when the intersection is empty, but only maximally}
    \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
    \If{$\Call{Dirty}{q} = \emptyset$}
    	\State \Return $\{ q \}$
   \ElsIf{$C_{min} = \emptyset$}
    	\State \Return $\{ \mu_{\Call{Dirty}{q}}(q), q \}$
    \Else
    \State \Return $\{\mu_{\Call{Dirty}{q}}(q), \mu_{C_{min}}(q), \delta_{C_{min}}(q)\}$
    \EndIf
    \EndFunction

@@

    \Require{$t$ is a table and $\phi$ is a filter predicate.}
    \Ensure{Returns a set of optimal query plans for scanning and filtering $t$, with distinct dirty sets.}
    \Function{OptFilter}{$t, \phi$}
    \todo{jose: I believe this was missing the sigma before}
    \State \Return $\Call{OptRel}{\{  \sigma_{\phi}(q) | q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}} \} }$
    \EndFunction

@@

    \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
    \Ensure{Returns a set of optimal query plans for all joins, with distinct dirty sets.}
    \Function{OptJoin}{$Q, \Psi$}
    \For{$size \in 1...|\Psi|$}
    	\State $S \gets subset(\Psi, size)$
	\For{$\psi \in S$}
		\State $S' \gets S \setminus \psi$
			\If{$joins(S', \psi)$} 
				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
				\State $t \gets \Call{Rels}{\psi} \setminus \Call{Rels}{S'}$
				\State $t_{plans} \gets Q(t)$
				\For {$l, r \in t_{plans} \times S'_{plans}$}
					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\psi}}$
					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\psi}}$
					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\psi} p_r | p_l \in P_l, p_r \in P_r \} }$}
				\EndFor
		\EndIf
	\EndFor
    \EndFor
    \State \Return{Q(\Call{Rels}{$\Phi$})}
    \EndFunction
  \end{algorithmic}
  \caption{Base algorithms for query planning with imputations.}
  \label{algo:plan-helpers}
\end{algorithm}

>>>>

%\begin{algorithm}
  \begin{algorithmic}
    \Require{$q$ is a query plan, $C_{req}$ is a set of attributes that must be imputed.}
    \Ensure{Returns a set of query plans such that $\Call{Dirty}{q'} \cap C_{req} = \emptyset$.}
    \Function{AddImpute}{$q, C_{req}$}
    \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
    \If{$\Call{Dirty}{q} = \emptyset$}
    	\State \Return $\{ q \}$
   \ElsIf{$C_{min} = \emptyset$}
    	\State \Return $\{ \mu_{\Call{Dirty}{q}}(q), q \}$
    \Else
    \State \Return $\{\mu_{\Call{Dirty}{q}}(q), \mu_{C_{min}}(q), \delta_{C_{min}}(q)\}$
    \EndIf
    \EndFunction

@@

    \Require{$t$ is a table and $\phi$ is a filter predicate.}
    \Ensure{Returns a set of optimal query plans for scanning and filtering $t$, with distinct dirty sets.}
    \Function{OptFilter}{$t, \phi$}
    \State \Return $\Call{OptRel}{\{  \sigma_{\phi}(q) | q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}} \} }$
    \EndFunction

@@

    \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
    \Ensure{Returns a map from dirty sets to optimal query plans involving all necessary joins}
    \Function{OptJoin}{$Q, \Psi$}
    \For{$size \in 1...|\Psi|$}
    	\State $S \gets subset(\Psi, size)$
	\For{$\psi \in S$}
		\State $S' \gets S \setminus \psi$
			\If{$joins(S', \psi)$} 
				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
				\State $t \gets \Call{Rels}{\psi} \setminus \Call{Rels}{S'}$
				\State $t_{plans} \gets Q(t)$
				\For {$l, r \in S'_{plans} \times t_{plans}$}
					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\psi}}$
					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\psi}}$
					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\psi} p_r | p_l \in P_l, p_r \in P_r \} }$}
				\EndFor
		\EndIf
	\EndFor
    \EndFor
    \State \Return{Q(\Call{Rels}{$\Phi$})}
    \EndFunction
  \end{algorithmic}
  \caption{Base algorithms for query planning with imputations.}
  \label{algo:plan-helpers}
%\end{algorithm}

### 2016-12-12 paper/examples.tex

Consider the case of an analyst looking to explore the sources and causes of
polling error in a presidential election. One hypothesis \cite{zukin2015s}
is that the rapid switch from landlines to cellphones has increased the
difficulty of contacting potential poll respondents. The analyst may
begin her analysis by comparing the polling error in a state with the
distribution of households with landlines, for the set of states for which
polling error was most pronounced. In order to probe deeply, the analyst
accesses a household-level survey of household characteristics that includes
whether the household has a landline. One such survey --- which will be
discussed further --- is the American Community Survey (ACS), conducted by the
US Census Bureau.

Surveys are rife with missing values due to non-response and other issues, and
the ACS is no exception. The missing values could be imputed in an expensive
operation\footnote{One would not necessarily want to impute the entire database once and
store the imputed values, as in many settings the imputation strategy would be
targeted for a specific question or domain.}, or the rows with missing values could be dropped entirely. With
ImputeDB, the analyst could impute the relevant subset of data on-the-fly,
achieving better performance without losing the information in the dropped
examples.

The analyst submits a query to ImputeDB which finds an optimized query plan.

\begin{verbatim}
SELECT polling.state, AVG(acs.TEL)
FROM polling, acs
WHERE polling.state = acs.ST
  AND polling.error > 50
GROUP BY polling.state
\end{verbatim}

\todo{explore a couple of plans generated, using dot visualizer}

>>>>

Consider the case of an analyst looking to explore the sources and causes of
polling error in a presidential election. One hypothesis \cite{zukin2015s}
is that the rapid switch from landlines to cellphones has increased the
difficulty of contacting potential poll respondents. The analyst may
begin her analysis by comparing the polling error in a state with the
distribution of households with landlines, for the set of states for which
polling error was most pronounced. In order to probe deeply, the analyst
accesses a household-level survey of household characteristics that includes
whether the household has a landline. One such survey --- which will be
discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
conducted by the US Census Bureau.

Surveys are rife with missing values due to non-response and other issues, and
the ACS is no exception. In this case, a non-negligible fraction of respondents omitted
information on whether they have a landline. The analyst hypothesizes that older houses are
more likely to have landlines and that households that are relatively new are less likely to
have landlines.

The analyst could take advantage of these presumed correlations to impute the missing values
in an expensive operation\footnote{One would not necessarily want to impute the entire
database once and store the imputed values, as in many settings the imputation strategy
would be targeted for a specific question or domain.}. On the other hand, the query
could be run directly and rows with missing values would be dropped entirely. With
ImputeDB, the analyst could impute the relevant subset of data on-the-fly, achieving
better performance without losing the information in the dropped examples.

The analyst submits a query to ImputeDB which finds an optimized query plan. The resulting
query imputes missing values for \verb!acs.TEL! after states with high polling error have
been selected.

\begin{verbatim}
SELECT polling.ST, AVG(acs.TEL)
FROM polling, acs
WHERE polling.ST = acs.ST
  AND polling.ERROR > 50         -- 5 percentage points
GROUP BY polling.ST
\end{verbatim}

\todo{Explore a couple of plans generated, using dot visualizer. Confirm that the actual
plans match my guess in last sentence.}

### 2016-12-12 paper/introduction.tex

In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
The guiding design principle of \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
To achieve this goal, we perform imputation on the fly, during query execution.
Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows our system to flexibly trade accuracy for computation time.

>>>>

In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
To achieve this goal, we perform imputation on the fly, during query execution.
Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade accuracy for computation time.

### 2016-12-12 paper/relatedwork.tex

\section{Related Work}
\subsection{Missing Values and Statistics}
Imputation of missing values is a widely studied field within the statistics and machine
learning communities. As highlighted in \cite{gelman2006data}, missing data
can appear for a variety of reasons, including both random and conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between the attributes to factor in
such varied forms of missingness.

Burgette and Reiter introduce imputation of missing data using sequences
of regression trees in \cite{burgette2010multiple}. \ProjectName{} uses
this technique to impute missing values as needed in the generated query plan.
The flexibility of this non-parametric imputation approach aligns well with the
spirit of query optimization, which aims to reconcile performance with a declarative interface to
data manipulation. By removing concerns for imputation, users executing queries
in \ProjectName{} don't need to consider the potential nature of the missing data.

In \cite{akande2015empirical}, Akande et al analyze the performance of various
imputation techniques on American Community Survey data, "an ongoing survey
that provides vital information on a yearly basis"\cite{censusbureau}. The authors
use a common technique: multiple imputation, which imputes missing values
with various strategies on the base data, computes aggregates, and then
combines estimates of these aggregates to account for additional uncertainty across
strategies. The computational difficulties of imputing on large base tables
are well known and can limit approaches. For example, Akande finds that
one of their approaches (MI-GLM) crashes when attempting to impute on data that
includes variables with potentially large domains (ten categories specifically).
 In contrast, \ProjectName{} allows users to specify a tradeoff
between information loss and computational complexity (in terms of time). 
Furthermore, the query planner's imputation is guided by the requirements of
each specific query's operators, rather than requiring broad assumptions
about query workloads.

>>>>

\section{Related Work}

\todo{Micah: Should this section follow the introduction? A lot of this stuff
becomes repetitive as we wouldn't be able to develop the body of the paper
without having already made these citations.}

\subsection{Missing Values and Statistics}
Imputation of missing values is a widely studied field within the statistics and machine
learning communities. As highlighted in \cite{gelman2006data}, missing data
can appear for a variety of reasons, including both random and conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between the attributes to factor in
varied forms of missingness.

Burgette and Reiter (\cite{burgette2010multiple}) discuss the usage of
sequential regression trees for imputing missing data.

In \cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset
(Sec~\ref{subsec:datasets}). The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approaches (MI-GLM) crashes when attempting to impute on data that
includes variables with potentially large domains (ten categories specifically).
In contrast, \ProjectName{} allows users to specify a tradeoff between
information loss and computational complexity (in terms of time). Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

### 2016-12-12 presentation/main.tex

\begin{frame}{Overview}
\begin{itemize}
	\item Handling incorrect or dirty data is a complex and challenging problem for data scientists
	\item $\text{missing data} \subset \text{dirty data}$
	\item Cost of running imputation on entire dataset
	\item Lack of flexibility
	\item \ProjectName{} allows users to interact with dataset as though it were clean.
	\item User should never see missing data or have to modify query.
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Query Planning: Overview}
\begin{itemize}
	\item $\mu_{min}, \mu_{max}, \delta$: new relational algebra operators
	\item Cost of computation vs. information loss: key tradeoff controlled by $\alpha$ parameter
	\item Placement of operators and dirty sets generated considered in planning.
	\item Histogram transformations used to provide up-to-date cardinality and missing value estimates during planning.
	\item Dynamic programming approach, extends System R optimization\cite{blasgen1981system}
	\item Exponential complexity, but tractable in practice with low planning times.
\end{itemize}
\end{frame}

\begin{algorithm}[H]
 \begin{algorithmic}
   \scriptsize
   \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
   \Ensure{Returns optimized query plan.}

  \Function{Plan}{$T, \Phi, \Psi, P, G, A$}
 \State  $Q \gets \{ \}$ \Comment{map from table and dirty set to plans }
   \For{$t \in T$}
   	\State $\phi \gets \Call{GetFilters}{t, \Phi}$
	\State $P_{t} \gets \Call{OptFilter}{t, \phi}$
	\State $Q \gets Q \cup \{ (t, \Call{Dirty}{q}) \mapsto q | q \in P_t \}$
   \EndFor

   \State $M \gets \{ \}$ \Comment{map from dirty set to plans, contains all joins necessary}
   \State $M \gets  \Call{OptJoin}{Q, \Psi}$	\Comment{optimize joins}
   \If{$G \neq \emptyset \land A \neq NULL$}
   	\State $attrs \gets \Call{Attrs}{G} \cup \Call{Attrs}{A}$
   	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \Call{GroupBy}{\Call{AddImpute}{q, attrs}, G, A}}$
   \Else
  	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \pi_{P}(\Call{AddImpute}{q, P}) }$
  \EndIf

\Return $\argmin_{q \in M} Cost(q)$
\EndFunction

  \end{algorithmic}
  \caption{Top-level query planner with imputations.}
\label{algo:top-level-planner}
\end{algorithm}

\begin{frame}[fragile]{Imputation Algorithm}
\begin{itemize}
	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CERT)\cite{burgette2010multiple}
	\item Significant cost to perform on entire dataset
	\item \ProjectName{} allows flexible use
\end{itemize}

\begin{algorithm}[H]
  \scriptsize
  \begin{algorithmic}
    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
    \Ensure{Returns an imputed $T$}

    \Function{ImputeWithCERT}{$H, op$}
    	\State $T' \gets \Call{ImputeRandom}{T, D}$
	\For{$n\ EPOCHS$}
		\For{$d \in D$}
			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
		\EndFor
	\EndFor
	\Return{T'}
	\EndFunction
  \end{algorithmic}
  \caption{An algorithm for chained imputation using regression trees}
  \label{algo:imputation-strategy}
\end{algorithm}
\end{frame}

@@

>>>>

\begin{frame}{Introduction and Motivation}
\begin{itemize}
	\item Handling incorrect or dirty data is a complex and challenging problem for data scientists
	\item $\text{missing data} \subset \text{dirty data}$ and requires imputation/dropping.
	\item Cost of running imputation on entire dataset is a high barrier for analysis.
	\item Lack of flexibility inherent in existing pre-processing approaches.
	\item User should never see missing data or have to modify query. Want a DBMS that 
		allow us to "pretend" data is perfect.
	\item \ProjectName{} allows users to interact with dataset as though it were clean, planning
	around necessary imputation/drops.
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Query Planning: Overview}
\begin{itemize}
	\item $\mu_{min}, \mu_{max}, \delta$: new relational algebra operators
	\item Cost of computation vs. information loss: key tradeoff controlled by $\alpha$ parameter
	\item Placement of operators and dirty sets generated considered in planning.
	\item Histogram transformations used to provide up-to-date cardinality and missing value estimates during planning.
	\item Dynamic programming approach, extends System R optimization\cite{blasgen1981system}
	\item Exponential complexity, but tractable in practice with low planning times.
	\item Imputation operator is a state-of-the-art regression-tree based method.
\end{itemize}
\end{frame}

@@

\begin{frame}
	\center
	Extra slides with details for those interested
\end{frame}

\begin{algorithm}[H]
\scriptsize
\input{../paper/algorithms/plan-toplevel.tex}
\end{algorithm}

\begin{frame}[fragile]{Imputation Algorithm}
\begin{itemize}
	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CERT)\cite{burgette2010multiple}
	\item Significant cost to perform on entire dataset
	\item \ProjectName{} allows flexible use
\end{itemize}
\scriptsize

\begin{algorithm}[H]
\scriptsize
\input{../paper/algorithms/cert.tex}
\end{algorithm}
\end{frame}
